---
title: 'Final Project 2: Reproducible Report on COVID19 Data'
author: "Lynn Perez"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
---



This project looks at data concerning cases and deaths related to COVID19 from January 2020 to September 2022. We will be looking at a comparison between the number of cases and deaths resulting from COVID19 in the U.S. and globally.

```{r import_data}
# install r packages
library(tidyverse)
library(lubridate)
library(stringr)

#read in data
url_in <- ("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/")
file_names <- c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```

```{r tidy_global_data}
#Organize the global data
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), 
               names_to = "date", 
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), 
               names_to = "date", 
               values_to = "deaths") %>%
  select(-c(Lat,Long))

#Join global cases and deaths
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date)) %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

#read in population data
uid_lookup_url <- "https://raw.github.com/CSSEGISandData/COVID-19//master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

#add population data and calculate cases/deaths per million and deaths per case
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key) %>%
  group_by(date) %>%
  summarize(global_cases = sum(cases), global_deaths = sum(deaths), global_population = sum(Population, na.rm=T)) %>%
  mutate("global_cases_per_mill" = global_cases*100000 / global_population) %>%
  mutate("global_deaths_per_mill" = global_deaths*100000 / global_population) %>%
  mutate("global_deaths_per_case" = global_deaths / global_cases) %>%
  ungroup() %>%
  select(date, global_cases, global_deaths, global_population, global_cases_per_mill, global_deaths_per_mill,global_deaths_per_case)
```

``` {r tidy_us_data}
#Organize the US data
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

#Join US cases and deaths datasets and calculate cases/deaths per million and deaths per case
US <- US_cases %>%
  full_join(US_deaths) %>%
  group_by(Country_Region, date) %>%
  summarize(US_cases = sum(cases), US_deaths = sum(deaths), US_population = sum(Population)) %>%
  mutate(US_cases_per_mill = US_cases*100000 / US_population) %>%
  mutate(US_deaths_per_mill = US_deaths*100000 / US_population) %>%
  mutate(US_deaths_per_case = US_deaths / US_cases) %>%
  ungroup() %>%
  select(date, US_cases, US_deaths, US_population, US_cases_per_mill, US_deaths_per_mill,US_deaths_per_case)

#Join the global and US datasets
US_vs_global <- US %>%
  full_join(global) %>%
  select(-c(US_population,global_population))
```

##Global & U.S. COVID19 Cases

``` {r us_vs_global_cases}
#Graph US cases per million vs global cases per million
US_vs_global %>%
  ggplot(aes(x = date, y = global_cases_per_mill)) +
  geom_line(aes(color = "Global Cases")) +
  geom_line(aes(y = US_cases_per_mill, color = "US Cases")) + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 Cases", 
       x="Date",
       y="Cases Per Million People",
       color=NULL) +
  scale_color_manual(values = c("blue","red"))
```

This graph shows us that the U.S. has had more cases per million people than the rest of the world. This does not take into account how cases were reported in the U.S. compared to the reporting system of other countries.


##Global & U.S. COVID19 Deaths

``` {r us_vs_global_deaths}
#Graph US deaths per million vs global deaths per million
US_vs_global %>%
  ggplot(aes(x = date, y = global_deaths_per_mill)) +
  geom_line(aes(color = "Global Deaths")) +
  geom_line(aes(y = US_deaths_per_mill, color = "US Deaths")) + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 Deaths", 
       x="Date",
       y="Deaths per Million People",
       color=NULL) +
  scale_color_manual(values = c("blue","red"))
```

This graph shows a similar result, that the U.S. has had more deaths per million people than the rest of the world. Again, there may be bias becasue this does not take into account how reporting varies from country to country.


##Global & U.S. COVID19 Deaths Per Case

``` {r us_vs_global_deaths_per_case}
#Graph US deaths per case vs global deaths per case with prediction lines for both
  us_mod <- lm(US_deaths_per_case ~ date, data = US_vs_global)
  global_mod <- lm(global_deaths_per_case ~ date, data = US_vs_global)
  
US_vs_global <- US_vs_global %>% 
  mutate(US_pred = predict(us_mod)) %>%
  mutate(global_pred = predict(global_mod))

US_vs_global %>%
  ggplot(aes(x = date, y = US_deaths_per_case)) +
  geom_line(aes(color = "US Deaths Per Case"), size=1.5) +
  geom_line(aes(y = US_pred, color = "Predicted US Deaths Per Case"), linetype = "dashed") +
  geom_line(aes(y = global_deaths_per_case, color = "Global Deaths"), size=1.5) + 
  geom_line(aes(y = global_pred, color = "Predicted Global Deaths Per Case"), linetype = "dashed") +
  theme(legend.position="bottom") +
  ylim(0,.125) +
  labs(title = "COVID19 Global & US Deaths Per Case", 
       x="Date", 
       y="Deaths Per Case", 
       colour = NULL) +
  scale_color_manual(values = c("blue1","blue2","red1","red2"))
```

This graph shows that while the U.S. has had higher cases and deaths per million citizens compared to the rest of the world, the U.S. has a lower amount of deaths per case. This could be again because of bias due to reporting systems or may be due to a combination of other factors including available resources and information, communication systems, etc. I also included prediction lines for both U.S. and global amount of deaths per case. It appears that the U.S. has been decreasing its amount of deaths per case at a faster rate than the rest of the world.